EGGs are adhesive!

### 2025-03-04 calco/main.tex
minima

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled, acyclic) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the former machinery at work, showing how the optimisation steps
can be rephrased as suitable applications of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, proofs have been moved to the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be two arrows admitting kernel pairs and suppose that the solid part of the three squares below is given.

	If the leftmost square is commutative, then there exists a a unique arrow $k_h: K_f \to K_g$ making the other two commutative.
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z}\]

		Moreover, the following hold true:
		\begin{enumerate}
			\item if the leftmost squareis a pullback, then also the other two are so;
			\item if $h$ is mono then $k_h$ is mono too.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

The previous result allows us to deduce the following lemma in an $\mathcal{M}$-adhesive context.
\todo{A questo punto la proposizione seguente possiamo anche buttarla via se serve recuperare spazio}
\noindent
\parbox{7.5cm}{
\begin{restatable}{proposition}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar[rr]_{g} & & D }}

@@

\begin{proposition}\label[proposition]{prop:mono} Let $\Sigma$ be an algebraic signature. Then it holds
	\begin{enumerate}
		\item an arrow $(h,k)$ in $\hyp_{\Sigma}$ is a mono if and only if $h$ and $k$ are injective functions;
		\item $\hyp_{\Sigma}$ is an adhesive category. 
	\end{enumerate}
\end{proposition}

@@

\subsection{Labeling hypergraphs with equivalence}

@@

\begin{definition}
	Let $\mathcal{G} = (E, V, C, s, t, q)$ be a hypergraph with equivalence and $(S, \pi_1, \pi_2)$ a kernel pair of $q^\star \circ s$.
	Then, $\mathcal{G}$ is an \emph{e-hypergraph} whenever $q^\star \circ t \circ \pi_1 = q^\star \circ t \circ \pi_2$.
	$\egg$ is the full subcategory of $\EqHyp$ whose objects are e-hypergraphs, and $I: \egg \to \EqHyp$ is the inclusion functor.
\end{definition}

@@

\section{Pros and cons of adhesive rewriting}
\label{rewriting}
The previous sections has shown how term graphs with (possibly) closed equivalence
can be described as a suitable $\mathcal{M},\mathcal{N}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Of course, being a purely graphical representation, there can be redundancy and possibly ambiguity,
so that e.g. a term such as $f(a,a)$ admits a few different representations as a term graphs, given below

\[f(a,a) \mbox{ con 2 a sharate, con 2 a nella stessa classe di equivalenze, e con 2 a separate}\]

Admittedly, the third one is illegal, i.e. it is not an object of $\EqTGs$ but not one of $\GEqTGs$: constants have no input, 
and nodes that are targets of edges with the same label and source the empty word must be equivalent,
and possibly be the same.
%
In general, note that a maximally shared representation of an EGG exists, and for our 
toy example is the left-most above. In general terms,
for an EGG $\mathcal{G}$, such a canonical form is the final object of the comma category 
$\mathcal{G}  \downarrow \GEqTGs$. 
This is a standard result in term graph rewriting, see e.g.~\cite{xxx}, and it holds also in the setting with equivalences.
In fact, if $\mathcal{G}$ has no empty nodes, the unique arrow from $\mathcal{G}$ to the terminal object
is injective on equivalence classes.
%
\todo{vedere se mettere qualcosa prima sugli aciclici}
The restriction to the full category $\GEqATGs$ of those objects whose underlying graph is \emph{acyclic} is 
straightforward, replicating all the considerations 
above and in the previous sections, even if we refrain from explicitly presenting it.

What about rewriting? The theory of $\mathcal{M},\mathcal{N}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$ and the match is an arrow in $\mathcal{N}$, then we can lift all the standard properties
of the DPO approach to the category at hand. Luckily, this is the structure of the rules that is advanced in~\cite{DetlefsNS05},
which is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity and the second
component is a regular mono, simply adding the equivalences of interest.
%
For example, the equivalence $x \div x = 1$, from the introductory example in~\cite{xxx}, 
can be modelled as the rule
\[\mbox{to draw DPO rule, x in the same equivalence class} \]

An immediate drawback of choosing the identity as the left-hand side is the possibilty to keep on repeating the same rewriting step.
However, this is easily forbidden adopting rules \emph{with negative application conditions}~\cite{xxx}, i.e. the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$. For EGGs and rules $L \leftarrow L \rightarrow R$, $n$ is simply the right-hand side.

@@

\emph{Future works.}
Our result opens two threads of research. The first is to check how much of the $\mathcal{M},\mathcal{N}$-adhesivity 
of EGGs can be used to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be span of regular monos, and as such they
perfectly fit the mold of rewriting on $\mathcal{M}-\mathcal{N}$adhesive categories. The restriction on matches 
though still need to be further investigated.
%
Also, it still needs to be investigated how parallelism and causality, the key features for DPO rewriting
on adhesive categories, can be exploited in the context of EGGs implementation. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hyper-graphs can be generalised to hierarchical hyper-graphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges that is useful for adding structural information, such as encapsulation and sandboxing~\cite{ghica, noi}.

>>>>

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled, acyclic) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the former machinery at work, showing how the optimisation steps
can be rephrased as suitable applications of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be two arrows admitting kernel pairs and suppose that the solid part of the three squares below is given. 
	%
	If the leftmost square is commutative, then there is a a unique arrow $k_h: K_f \to K_g$ making the other two commutative
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z}\]

		Moreover, it holds
		\begin{enumerate}
			\item if the leftmost square is a pullback, then also the other two are so;
			\item if $h$ is mono then $k_h$ is mono too.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

The previous lemma allows us to deduce the following results in an $\mathcal{M}$-adhesive context.
\todo{A questo punto la proposizione seguente possiamo anche buttarla via se serve recuperare spazio}
\noindent
\parbox{7.5cm}{
\begin{restatable}{proposition}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar[rr]_{g} & & D }}

@@

\begin{proposition}\label[proposition]{prop:mono} Let $\Sigma$ be an algebraic signature. Then it holds
	\begin{enumerate}
		\item an arrow $(h,k)$ in $\hyp_{\Sigma}$ is a mono if and only if $h$ and $k$ are injective;
		\item $\hyp_{\Sigma}$ is an adhesive category. 
	\end{enumerate}
\end{proposition}

@@

\subsection{Term graphs with equivalence}

@@

\subsection{Hypergraph with closed equivalence}
\begin{definition}
	Let $\mathcal{G} = (E, V, C, s, t, q)$ be a hypergraph with equivalence and $(S, \pi_1, \pi_2)$ a kernel pair of $q^\star \circ s$.
	Then, $\mathcal{G}$ is an \emph{e-hypergraph} whenever $q^\star \circ t \circ \pi_1 = q^\star \circ t \circ \pi_2$.
	$\egg$ is the full subcategory of $\EqHyp$ whose objects are e-hypergraphs, and $I: \egg \to \EqHyp$ is the inclusion functor.
\end{definition}

@@

\subsection{Term graphs with closed equivalence}
\todo{Here we must introduce $\GEqTGs$}

\section{Pros and cons of adhesive rewriting}
\label{rewriting}
The previous sections has shown how term graphs with equivalence,
possibly closed with respect to operator composition,
can be described as a suitable $\mathcal{M},\mathcal{N}$-adhesive category. 
The latter precisely capture EGGS, as originally
presented in~\cite{WillseyNWFTP21}. 
%
Being a purely graphical representation, EGGS allows for redundancy and possibly ambiguity,
so that e.g. a term such as $a \div a$ admits a few different representations as a term graph, as shown below
%
\[a \div a \mbox{ con 2 a sharate, con 2 a nella stessa classe di equivalenze, e con 2 a separate}\]
%
Admittedly, the third one is illegal, i.e. it is an object of $\EqTGs$ but not one of $\GEqTGs$: constants have no input, 
and nodes that are targets of edges with the same label and source the empty word must be equivalent,
and possibly the same.
%
In general, note that a maximally shared representation of an EGG exists, and for our 
toy example is the left-most above. In general terms,
for an EGG $\mathcal{G}$, such a canonical form is the final object of the comma category 
$\mathcal{G}  \downarrow \GEqTGs$. 
This is a standard result in term graph rewriting, see e.g.~\cite{xxx}, and it holds also in the setting with equivalences.
In fact, if $\mathcal{G}$ has no empty nodes, the unique arrow from $\mathcal{G}$ to the terminal object
is injective on equivalence classes.
%
\todo{vedere se mettere qualcosa prima sugli aciclici}
The restriction to the full category $\GEqATGs$ of those objects whose underlying graph is \emph{acyclic} is 
straightforward, replicating all the considerations 
above and in the previous sections.
% even if we refrain from explicitly presenting it.

What about rewriting? The theory of $\mathcal{M},\mathcal{N}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$ and the match is an arrow in $\mathcal{N}$, then we can lift all the standard properties
of the DPO approach to the category at hand. Luckily, this is the structure of the rules that is advanced in~\cite{DetlefsNS05},
which is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity and the second
component is a regular mono, simply adding the edges, nodes, and equivalence classes of interest.
%
An EGG rule such $x \div x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}, 
can be modelled as the DPO rule below,
which can be applied with a match that is injective on equivalence classes to the left-most and the middle EGG
in the picture above
\[\mbox{to draw DPO rule, x in the same equivalence class} \]

An immediate drawback of choosing the identity as the left-hand side is the possibility to keep on repeating the same rewriting step.
However, this is easily forbidden adopting rules \emph{with negative application conditions}~\cite{xxx}, which are given by the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$. For EGGs and rules $L \leftarrow L \rightarrow R$, $n$ is simply the right-hand side itself.
%
The theory of $\mathcal{M},\mathcal{N}$-adhesivity carries on with such enriched rules, see e.g~\cite{xxx}.

Summing up, showing that EGGs are objects of $\GEqTGs$ and that optimisation steps are obtained via 
DPO rules with negative application conditions unleash the xxx.

@@

\emph{Future works.}
Our result on $\GEqTGs$ opens a few threads of research. The first is to check how the $\mathcal{M},\mathcal{N}$-adhesivity 
of EGGs can be pushed to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be span of regular monos (possibly with negative application 
conditions), and as such they perfectly fit the mould of rewriting on $\mathcal{M}, \mathcal{N}$-adhesive categories. 
The consequences for usability on the restriction of the matches to arrows that are injective on equivalence classes
though need to be further investigated.
%
Also, it still needs to be exploited how parallelism and causality, the key features for DPO rewriting
on adhesive categories, can be exploited in the context of EGGs implementation. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hyper-graphs can be generalised to hierarchical hyper-graphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges~\cite{ghicaZan,CastelnovoGM24}, which is useful for adding structural information, such as encapsulation and sandboxing.
